---
title: "Bank Marketing Data - A Decision Tree Approach"
authors: Tarlan Aghayev, Wojciech Misiura
date: "2024-05-03"
output:
  html_notebook: default
  html_document: default
---


Reference link to kaggle <https://www.kaggle.com/code/shirantha/bank-marketing-data-a-decision-tree-approach>

```{r}

```

## Context

As part of Reproducible Research course which is Mandatory course for 2nd year students of Data Science and Business Analytics, we are reproducing already existing project.

Link to the project: <https://www.kaggle.com/code/shirantha/bank-marketing-data-a-decision-tree-approach>

The aim of this attempt is to

```{r, message=FALSE, warning=FALSE, echo=FALSE}
#install.packages("remotes")
#library(remotes)

packages_version <- c(
  "readr" = "2.1.4", 
  "dplyr" = "1.1.2", 
  "ggplot2" = "3.4.2",
  "fastDummies" = "1.7.3",
  "rpart" = "4.1.23",
  "caret" = "6.0.94",
  "rpart.plot" = "3.1.2",
  "pROC" = "1.18.5",
  "corrplot" = "0.92",
  "RColorBrewer" = "1.1.3",
  "tibble" = "3.2.1",
  "rpart" = "4.1.23")

check_and_install <- function(package, version) {
  if (!require(package, character.only = TRUE)) {
    package_version_install <- paste0(package, "@", version)
    remotes::install_version(package_version_install)
  }
}


mapply(check_and_install, names(packages_version), packages_version)
```



```{r}
url<-"https://raw.githubusercontent.com/WojciechMisiura/RRproject/main/bank.csv" 

bank<-read_csv(url)
```

```{r}
bank %>% summarize(across(everything(), ~sum(is.na(.))))

```

<<<<<<< HEAD
```{r}
summary(bank)
```


```{r}
ggplot(bank, aes(y = age)) + 
  geom_boxplot() + 
  theme_minimal() + 
  labs(title = "Boxplot for 'age'")
```

```{r}
ggplot(bank, aes(x = age)) +
  geom_histogram(bins = 100, fill = "blue", color = "black") +
  labs(title = "Distribution of Age") +
  theme_minimal()
```

```{r}
ggplot(bank, aes(y = duration)) + 
  geom_boxplot() + 
  theme_minimal() + 
  labs(title = "Boxplot for 'duration'")
```

```{r}
ggplot(bank, aes(x = duration)) +
  geom_histogram(bins = 100, fill = "blue", color = "black") +
  labs(title = "Distribution of Duration") +
  theme_minimal()
```

# Convert categorical variables

```{r}
bank_data <- bank
```

#### ------------------------------ job ------------------------------

```{r}
jobs <- c('management', 'blue-collar', 'technician', 'admin.', 'services', 
          'retired', 'self-employed', 'student', 'unemployed', 'entrepreneur')
```

## Explatory Data Analysis

```{r}
# Create a boxplot for 'age'
g_age <- ggplot(bank, aes(x = factor(0), y = age)) + 
  geom_boxplot() +
  labs(x = NULL, y = "Age") +
  theme(axis.title.x = element_blank())
g_age

# Create a distribution plot for 'age'
g_age_dist <- ggplot(bank, aes(x = age)) + 
  geom_histogram(bins = 100, fill = "blue", color = "black") +
  ggtitle("Distribution of Age")
g_age_dist

# Create a boxplot for 'duration'
g_duration <- ggplot(bank, aes(x = factor(0), y = duration)) + 
  geom_boxplot() +
  labs(x = NULL, y = "Duration") +
  theme(axis.title.x = element_blank())
g_duration

# Create a distribution plot for 'duration'
g_duration_dist <- ggplot(bank, aes(x = duration)) + 
  geom_histogram(bins = 100, fill = "red", color = "black") +
  ggtitle("Distribution of Duration")
g_duration_dist

```

## Convert Categorical Data

```{r}
# Copy data for processing
bank_data <- bank

# Count of people who made a deposit by job category
jobs <- c('management', 'blue-collar', 'technician', 'admin.', 'services',
          'retired', 'self-employed', 'student', 'unemployed', 'entrepreneur','housemaid', 'unknown')

for (j in jobs) {
  count <- sum(bank_data$deposit == "yes" & bank_data$job == j)
  cat(sprintf("%-15s : %5d\n", j, count))}
```

```{r}
table(bank_data$job)
```

```{r}
bank_data$job[bank_data$job %in% c('management', 'admin.')] <- 'white-collar'
bank_data$job[bank_data$job %in% c('services', 'housemaid')] <- 'pink-collar'
bank_data$job[bank_data$job %in% c('retired', 'student', 'unemployed', 'unknown')] <- 'other'

```

```{r}
table(bank_data$job)
```

#### ------------------------------ poutcome ------------------------------

```{r}
bank_data$poutcome <- as.character(bank_data$poutcome)
bank_data$poutcome[bank_data$poutcome == "other"] <- "unknown"
table(bank_data$poutcome)

```

#### ------------------------------ contact ------------------------------


```{r}
bank_data <- select(bank_data, -contact)
```

#### ------------------------------ default and housing and loan ------------------------------

```{r}
bank_data$default_cat <- as.integer(bank_data$default == "yes")
bank_data$housing_cat <- as.integer(bank_data$housing == "yes")
bank_data$loan_cat <- as.integer(bank_data$loan == "yes")

bank_data <- select(bank_data, -c(default, housing, loan))
```

#### ------------------------------ month, day ------------------------------

```{r}
bank_data <- select(bank_data, -c(month, day))
```

#### ------------------------------ deposit ------------------------------

```{r}
bank_data$deposit_cat <- as.integer(bank_data$deposit == "yes")
bank_data <- select(bank_data, -deposit)
```

#### ------------------------------ pdays ------------------------------

```{r}
cat("Customers that have not been contacted before:", sum(bank_data$pdays == -1), "\n")
cat("Maximum values on pdays:", max(bank_data$pdays), "\n")

bank_data$pdays[bank_data$pdays == -1] <- 10000

bank_data$recent_pdays <- ifelse(bank_data$pdays > 0, 1/bank_data$pdays, 1/bank_data$pdays)
bank_data <- select(bank_data, -pdays)
```

```{r}
tail(bank_data)
```

### ------------------------------ Convert to dummy values ------------------------------

```{r}
bank_with_dummies <- dummy_cols(bank_data, select_columns = c('job', 'marital', 'education', 'poutcome'),
                                remove_selected_columns = TRUE)
head(bank_with_dummies)
```

```{r}
print(dim(bank_with_dummies))

summary(bank_with_dummies)
```
### Observations on whole population

```{r}
ggplot(bank_with_dummies, aes(x = age, y = balance)) +
  geom_point(alpha = 0.5) +
  labs(title = "Scatterplot of Age vs. Balance",
       subtitle = "Across all ages, majority of people have savings of less than 20000",
       x = "Age", y = "Balance") +
  theme_minimal()
```

```{r}
ggplot(bank_with_dummies, aes(x = duration)) + 
  geom_histogram(data = subset(bank_with_dummies, poutcome_success == 1), fill = "blue", bins = 30) +
  labs(title = "Histogram of Duration for 'poutcome_success'",
       x = "Duration", y = "Count") +
  theme_minimal()
```

#### Analysis on people who sign up for a term deposite

```{r}
bank_with_dummies %>% 
  filter(deposit_cat == 1) %>%
  summary()
```

```{r}
sum(bank_with_dummies$deposit_cat == 1 & bank_with_dummies$loan_cat == 1 & bank_with_dummies$housing_cat == 1)
```

```{r}
sum(bank_with_dummies$deposit_cat == 1 & bank_with_dummies$default_cat == 1)
```

```{r}
ggplot(data = bank_data, aes(x = job, y = deposit_cat, fill = job)) +
  geom_bar(stat = "summary", fun = mean) +
  labs(title = "Average Deposit Category by Job", x = "Job Category", y = "Average Deposit Category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```


```{r}
ggplot(data = bank_data, aes(x = poutcome, y = duration, fill = poutcome)) +
  geom_bar(stat = "summary", fun = mean) +
  labs(title = "Average Duration by Outcome", x = "Outcome", y = "Average Duration") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  
```

> ## Classification

```{r}
bankcl <- bank_with_dummies
```

```{r}
corr <- cor(bankcl, use = "complete.obs") 
print(corr)
```

```{r}
corrplot(corr, method = "color", type = "upper", order = "hclust",
         tl.col = "black", tl.srt = 45,
         col = colorRampPalette(c(brewer.pal(10, "RdBu")[10:1]))(200),
         title = "Heatmap of Correlation Matrix", 
         cl.lim = c(-0.3, 0.3))
```

```{r}
corr['deposit_cat']
typeof(corr)

corr_deposit <- as.data.frame(t(corr_row))
names(corr_deposit) <- "deposit_cat"  

corr_deposit$variable <- rownames(corr_deposit)
rownames(corr_deposit) <- NULL 

corr_deposit <- arrange(corr_deposit, desc(deposit_cat))
print(corr_deposit)
```

> ## Build the Data Model

```{r}
set.seed(50) 

label <- bankcl$deposit_cat
data_drop_deposite <- bankcl[, !names(bankcl) %in% 'deposit_cat']

index_train <- createDataPartition(label, p = 0.8, list = FALSE)
data_train <- data_drop_deposite[index_train, ]
label_train <- label[index_train]
data_test <- data_drop_deposite[-index_train, ]
label_test <- label[-index_train]
```

```{r}
train_and_evaluate <- function(depth) {
  model <- rpart(formula = deposit_cat ~ ., data = bankcl, method = "class",
                 control = rpart.control(maxdepth = depth, usesurrogate = 0, xval = 10))
  train_pred <- predict(model, data_train, type = "class")
  test_pred <- predict(model, data_test, type = "class")
  train_accuracy <- sum(train_pred == label_train) / length(label_train)
  test_accuracy <- sum(test_pred == label_test) / length(label_test)

  cat(sprintf("Depth %d: Training score: %f, Testing score: %f\n", depth, train_accuracy, test_accuracy))
  return(c(train = train_accuracy, test = test_accuracy)) 
}


```

#### Compare Training and Testing scores for various tree depths used

```{r}
scores <- list()


depth_levels <- c(1, 2, 3, 4, 6)
names(depth_levels) <- c("max", "2", "3", "4", "6") 

for (depth in depth_levels) {
    result <- train_and_evaluate(depth)
    scores[[as.character(depth)]] <- list(train = result[1], test = result[2])
}

cat(sprintf('%-10s %-20s %-20s\n', 'depth', 'Training score', 'Testing score'))
cat(sprintf('%-10s %-20s %-20s\n', '-----', '--------------', '-------------'))
for (depth in c("2", "3", "4", "6", "max")) {
    cat(sprintf('%-1s %-25.2f %-20.2f\n', depth, scores[[depth]]$train, scores[[depth]]$test))
}
```

```{r}
dt2 <- rpart(label_train ~ ., data = data_train, method = "class", control = rpart.control(maxdepth = 2))

fi <- dt2$variable.importance

features <- names(fi)
l <- length(features)
for (i in 1:l) {
  cat(sprintf('%-20s %3f\n', features[i], fi[i]))
}
```

# Prediction



```{r}
cat(sprintf("Mean duration   : %f\n", mean(data_drop_deposite$duration, na.rm = TRUE)))
cat(sprintf("Maximum duration: %d\n", max(data_drop_deposite$duration, na.rm = TRUE)))
cat(sprintf("Minimum duration: %d\n", min(data_drop_deposite$duration, na.rm = TRUE)))
```

```{r}
new_data_371 <- data.frame(matrix(0, nrow = 1, ncol = length(names(data_drop_deposite))))
names(new_data_371) <- names(data_drop_deposite)
new_data_371$duration <- 371  # Set duration for prediction

new_data_3881 <- data.frame(matrix(0, nrow = 1, ncol = length(names(data_drop_deposite))))
names(new_data_3881) <- names(data_drop_deposite)
new_data_3881$duration <- 3881  # Set duration for prediction


probabilities_371 <- predict(dt2, new_data_371, type = "prob")
probabilities_3881 <- predict(dt2, new_data_3881, type = "prob")

class_371 <- predict(dt2, new_data_371, type = "class")
class_3881 <- predict(dt2, new_data_3881, type = "class")

cat("Probabilities with duration 371 sec:\n")
print(probabilities_371)

cat("Predicted class with duration 371 sec:\n")
print(class_371)

cat("Probabilities with duration 3881 sec:\n")
print(probabilities_3881)

cat("Predicted class with duration 3881 sec:\n")
print(class_3881)
```

```{r}
specific_row <- data_drop_deposite[986, ]
```


```{r}
probability_prediction <- predict(dt2, specific_row, type = "prob")

print(probability_prediction)
```

```{r}
preds <- predict(dt2, data_test, type = "class")

levels_set <- union(levels(factor(label_test)), levels(factor(preds)))

preds_factor <- factor(preds, levels = levels_set)
label_test_factor <- factor(label_test, levels = levels_set)

accuracy_matrix <- confusionMatrix(preds_factor, label_test_factor)
accuracy <- accuracy_matrix$overall['Accuracy']

cat("\nAccuracy score: \n", accuracy)
```

```{r}
probs <- predict(dt2, data_test, type = "prob")


positive_probs <- probs[, 2]

auc_value <- roc(label_test, positive_probs)$auc
cat("\nArea Under Curve: \n", auc_value)
```

```{r}

```

```{r}

```

```{r}

```
